---
title: "What is the Utility of Estimating Latent vs. Observed Change?"
subtitle: "Seminar on Research Designs"
author: Andreas M. Brandmaier
format: 
  revealjs:
    theme: [simple, style.css] #  “simple”, “dark”, “black”, “sky”, “beige”, “serif”, “solarized”, “blood”, “moon”, “night”, “league”, or  “white”
    footer: "6th ESLR Workshop, 2023: Social Learning in Complex Systems"
    highlight-style: a11y
    revealjs-plugins:
    - countdown
editor: visual
date: "05/07/2023"
editor: visual
---

```{r init, echo=FALSE}
library(tidyverse)
```

## Why is change interesting?

::: callout-tip
## "When you're finished changing, you're finished."

\- Benjamin Franklin
:::

## Why is change interesting?

--

# Difference scores

- Classic within-person research questions in educational and psychological research concerns a comparison of pre- and post-scores
- Difference scores are simply the difference between two scores

# The (un)reliability issue

Assume, we have two scores $X$ and $Y$ and form a differences score $D=X-Y$, the reliability of the difference score is:

$$\rho_D = \frac{\rho_{XX'}+\rho_{YY'}-2\rho_{XY}}{2-2\rho_{XY}} $$

with

$\rho_{XY}$ 


## Common pitfalls

Regression to the mean

```{r}
N <- 2000
set.seed(1234)
x1 <- rnorm(n=N, mean=100, sd = 15)
x2 <- rnorm(n=N, mean=100, sd = 15)
tibble(x1,x2)%>%ggplot(aes(x=x1,y=x2))+geom_point()

ids <- which(x1 <= 70)

tibble(x1,x2) %>% filter(x1<=70) %>%
  ggplot(aes(x=x2))+geom_density()
```
